Global search algorithms for one-dimensional multiextremal optimization problems with nonlinear constraints

### 2017-10-03 2017/ParOptBook/Chapter4.tex
Chapter 6 is ready

\section{Problem statement}
In the present chapter, the one-dimensional problems of the constrained optimization will
be considered. Let $\varphi(x)$ and $g_j(x)\le 0,1\le j\le m$ be the real functions defined in a closed interval $[a,b]$ of the real axis. It is requires to find the values $x^*$ satisfying the condition
\begin{equation}
  \label{eq4:problem}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b], g_j(x)\le 0,1\le j\le m\}
\end{equation}

@@

\begin{table}
  \caption{}
\begin{center}
  \label{tab4:exp_results}
  \begin{tabular}{| l | c | c | c| }
    \hline
    & $k_1$ & $k_2$ & $k_3$ \\ \hline
    Scanning a uniform grid & 160000 & 90280 & 56476 \\ \hline
    Penalty Function Method &  375 & 375 & 375 \\ \hline
    Index method & 63 & 49 & 35 \\ \hline
  \end{tabular}
\end{center}
\end{table}

Here $k_i$ is the number of computations of the $i$ problem function, i.e., of the function $g_i(x )$.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_6.png}
  \caption{The minimization of an ''index'' function}
  \label{fig:4_6}
\end{figure}

\section{Index algorithm taking into account the existence of $\epsilon$-reserved solutions}
Let us return to the consideration of a one-dimensional problem of the kind \eqref{eq4:problem}
\begin{displaymath}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b], g_j(x)\le 0,1\le j\le m\}
\end{displaymath}
In the case when the objective function and the left-side parts of the constraints are partially computable, it is necessary to use the form \eqref{eq4:problem2}.
\example
{
Let us consider problem \eqref{eq4:problem} with $x\in[0,6,2,2]$, $m=3$ as an illustration:
\begin{gather*}
\varphi(x)=\cos(18x-3)\sin(10x-7)+1, \\
g_1(x)=\exp(-x/2)\sin(6x-1.5) \\
g_2(x)=\sin(4x-2.2)+\cos(6x-2.9) \\
g_3(x)=|x|\sin(2\pi x - 0.5).
\end{gather*}
}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_7.jpg}
  \caption{The solution of a problem by Index Method}
  \label{fig:4_7}
\end{figure}
The functions $g_j , 1\le j\le 4$, are pictured in Fig. \ref{fig:4_7}. Assuming a partial calculability, the arcs of
the functions corresponding to the domains $Q_j ,1\le j\le 4$, from \eqref{eq4:q} are presented in Fig. \ref{fig:4_8}.

@@

\section{Index method with adaptive order of checking for constraints}
According to the rules of the index method considered in the preceding subsections, every trial includes the successive check of the fulfillment of the problem constraints in the corresponding point of the search domain. The finding of the first violated constraint terminates the trial and initiates the transition to the next iteration. If a numeration of the constraints in the increasing order by the computational costs of their checking is introduced (i.e., if the simpler constraints are checked first) one can reduce essentially the overall computational costs in the course of optimization. The above is important for the problems, in which the estimating of the function value requires a considerable computational resources connected with a necessity of the numerical modelling of the behavior of the systems described by these functions (as it takes place, for example, in numerous problems of the optimal design).

@@

For many applied problems, it is naturally to assume that a constraint violated at a point in the search domain, will be violated in some neighborhood of this point as well. In this case, it might be useful to begin the next trials in the points of this neighborhood with checking the fulfillment of this particular constraint. In the scope of the above, it seems reasonable to design a global optimization algorithm allowing the modification of the constraint  check order taking into account the information obtained during the search. The check order design will be oriented onto the constraint violated at an iteration point to be found as early as possible.

@@

Assume the constraint with the number $j,\: 1\le j\le m$, to fulfill at all points of the domain
\begin{equation}
  Q_j=\{x\in[a,b]:g_j(x)\le 0\},\: 1\le j\le m,
\end{equation}
which hereinafter will be called the feasible domain for this constraint. The feasible domain $Q$ of initial problem \eqref{eq4:problem} can be defined by the condition
\begin{equation}
  \label{eq4:Q_union}
  Q=Q_1\cap\dots\cap Q_m.
\end{equation}

@@

\begin{table}
  \caption{The number of problems with a small measure of the interval $[\alpha,\beta]$}
\begin{center}
  \label{tab4:exp_results2}
  \begin{tabular}{| c | c | c | c| c | c | c | c | c| c |}
    \hline
    \% & 0.01 & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ \hline
    $G_{SH}$ & 0 & 1 & 2 & 1 & 0 & 0 & 1 & 1 & 1  \\ \hline
    $G_{HL}$ & 3 & 6 & 2 & 3 & 5 & 1 & 2 & 1 & 2 \\ \hline
  \end{tabular}
\end{center}
\end{table}

@@

The first $p$ trial points are selected at the internal points of the interval $[a,b]$ arbitrarily. Each processor starts the trials execution at the corresponding points. After completing the $k^{th}$ trial, when the points $x^{k+1},\dots, x^{k+p - 1}$ have already been selected but the trial results at these points are not obtained yet, the point of the next trial $x^{k+p}$ is selected according to the following rules.

@@

\subsection{The convergence conditions of the algorithm}
\begin{theorem}
  \label{th4:convergence_par}
  Assume problem \eqref{eq4:problem} to have a solution $x^*$ and the following conditions to be fulfilled:
  \begin{itemize}
    \item each subdomain $Q_j ,1\le j\le m+1$ from \eqref{eq4:q} is a union of a finite number of positive length segments;
    \item each function $g_j ,1\le j\le m+1$ satisfies Lipschitz condition with the corresponding constant $L_j$;
    \item the components $\epsilon_\nu$ of the reserve vector $\epsilon_R$ satisfy the inequalities
    \begin{equation}
      \label{eq4:reserves_par}
      0<2\epsilon_\nu<L_\nu(\beta-\alpha)
    \end{equation}
    where $\beta-\alpha$ is the length of the interval $[\alpha,\beta]$ belonging to the feasible domain $Q$ from \eqref{eq4:Q_union} and including the point $x^*$ which is the solution of the problem;
    \item beginning from sufficiently large step number $k$, the values $\nu_\nu$ from \eqref{eq4:lip_est_par} corresponding to the nonempty sets $I_\nu$ satisfy the inequalities
    \begin{equation}
      r_\nu\mu_\nu>2L_\nu,\:1\le\nu\le m+1,
    \end{equation}
    where the algorithm parameters $r_\nu>1,\:1\le\nu\le m+1$;
    \item during the parallelization, the inequalities
    \begin{displaymath}
      T^j_{l+1}-T^j_j<T<\infty,1\le j\le p, l>1,
    \end{displaymath}
    are true where $T_l^j$ is the moment of completing the $l$ th trial by the $j^{th}$ processor and the inequality $1<p<P<\infty$ takes place for the number of processors $p$.
  \end{itemize}
  Then:
  \begin{itemize}
    \item the point $x^*$ is a limit one for the sequence ${x^k }$ generated by the described asynchronous algorithm for problem \eqref{eq4:problem} if $\epsilon=0$ in termination condition \eqref{eq4:stop_cond_par};
    \item any limit point $\overline x$ of the sequence $\{x_k \}$ is a solution of the problem \eqref{eq4:problem};
    \item the convergence to the limit point $\overline x$ is bilateral if $\overline x\not=a$ and $\overline x\not=b$.
  \end{itemize}
\end{theorem}

@@

\begin{proof}
  Since $\overline x$ is a limit point, there exists a sequence of the trial numbers $q^1,q^2,\dots$ satisfying the conditions
  \begin{equation}
    x^{q+1},\overline x\in[x_{t-1},x_t], t=t(q),q\in\{q^p\},p=1,2,\dots.
  \end{equation}
  These conditions reflect the fact that the point of the $(q+1)^{th}$ trial falls into the interval $[x_{t-1},x_t]$ containing the point $\overline x$ at the step $q$, i.e., $t=t(q)$.

@@

  At every step $k\ge 1$ there exists an interval $(x_{i-1},x_i), \:i=i(k)$, at one of the ends of which the value $z^*_\omega$ from \eqref{eq4:z_const_par} is achieved. This value corresponds to the index $\omega$ defined (when $\omega<m$) by the conditions $I_\omega\not=\O,\: I_{\omega+1}\not=\O$. If $I_{m+1}=\O$, the index $\omega=m+1$. If at that $\nu(x_{i -1} )\not = \nu(x_i )$, then according (4.47) and (4.50), $R(i)=2\Delta_i>0$. In the case when $\nu(x_{i -1} )= \nu(x_i )=\omega$, the estimate
  \begin{displaymath}
    \max(z_{i-1}, z_i)-z^*_\omega\le\mu\Delta_i
  \end{displaymath}
  follows from \eqref{eq4:lip_est_par} and, along with \eqref{eq4:characteristic_par}, gives the estimate
  \begin{displaymath}
    R(i)=\Delta_i\left(1-\frac{\max(z_{i-1},z_i)-z^*_\omega}{r_\omega\mu_\omega\Delta_i}\right)\ge (1-r^{-1})\Delta_i,
  \end{displaymath}
  where $r$ is from (4.61). Thus, at any search step $k\ge 1$ there exists an interval having a strictly positive characteristic. Hence, taking into account the rules of selection of the interval with the greatest characteristic \eqref{eq4:MaxR_par}, one can conclude that the intervals from system (4.54) should also have the positive characteristics since these ones undergo splitting up by the trial points falling into these intervals. Consequently, conditions (4.57) hold.

>>>>

\section{Problem statement}
In the present chapter, the one-dimensional problems of the constrained optimization will
be considered. Let $\varphi(x)$ and $g_j(x)\le 0,1\le j\le m$ be the real functions defined in a closed interval $[a,b]$ of the real axis. It is requires to find the values $x^*$ satisfying the condition
\begin{equation}
  \label{eq4:problem}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b], g_j(x)\le 0,\; 1\le j\le m\}
\end{equation}

@@

\begin{table}
  \caption{The results of solving the test problem}
\begin{center}
  \label{tab4:exp_results}
  \begin{tabular}{cccc}
    \hline
    Method & $k_1$ & $k_2$ & $k_3$ \\ \hline
    Uniform grid technique  & 160000 & 90280 & 56476 \\ 
    Penalty Function Method &  375 & 375 & 375 \\ 
    Index method & 63 & 49 & 35 \\ \hline
  \end{tabular}
\end{center}
\end{table}

Here $k_i$ is the number of computations of the $i$ problem function, i.e., of the function $g_i(x )$.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_6.png}
  \caption{The minimization of an ``index'' function}
  \label{fig:4_6}
\end{figure}

\section{Index algorithm taking into account the existence of $\epsilon$-reserved solutions}
Let us return to the consideration of a one-dimensional problem of the kind \eqref{eq4:problem}
\begin{displaymath}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b],\; g_j(x)\le 0, \; 1\le j\le m\}
\end{displaymath}
In the case when the objective function and the left-side parts of the constraints are partially computable, it is necessary to use the form \eqref{eq4:problem2}.
\example
{
Let us consider problem \eqref{eq4:problem} with $x\in[0.6,2.2]$, $m=3$ as an illustration:
\begin{gather*}
\varphi(x)=\cos(18x-3)\sin(10x-7)+1, \\
g_1(x)=\exp(-x/2)\sin(6x-1.5) \\
g_2(x)=\sin(4x-2.2)+\cos(6x-2.9) \\
g_3(x)=|x|\sin(2\pi x - 0.5).
\end{gather*}
}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_7.jpg}
  \caption{The solution of a problem by Index Method}
  \label{fig:4_7}
\end{figure}
The functions $g_j , 1\le j\le 4$, are pictured in Fig. \ref{fig:4_7}. Assuming a partial calculability, the arcs of
the functions corresponding to the domains $Q_j ,1\le j\le 4$, from \eqref{eq4:q} are presented in Fig. \ref{fig:4_8}.

@@

\section{Index method with adaptive order of checking for constraints}
According to the rules of the index method considered in the preceding subsections, every trial includes the successive check of the fulfilment of the problem constraints in the corresponding point of the search domain. The finding of the first violated constraint terminates the trial and initiates the transition to the next iteration. If a numeration of the constraints in the increasing order by the computational costs of their checking is introduced (i.e., if the simpler constraints are checked first) one can reduce essentially the overall computational costs in the course of optimization. The above is important for the problems, in which the estimating of the function value requires a considerable computational resources connected with a necessity of the numerical modelling of the behavior of the systems described by these functions (as it takes place, for example, in numerous problems of the optimal design).

@@

For many applied problems, it is naturally to assume that a constraint violated at a point in the search domain, will be violated in some neighbourhood of this point as well. In this case, it might be useful to begin the next trials in the points of this neighbourhood with checking the fulfilment of this particular constraint. In the scope of the above, it seems reasonable to design a global optimization algorithm allowing the modification of the constraint  check order taking into account the information obtained during the search. The check order design will be oriented onto the constraint violated at an iteration point to be found as early as possible.

@@

Assume the constraint with the number $j,\: 1\le j\le m$, is fulfilled at all points of the domain
\begin{equation}
  Q_j=\{x\in[a,b]:g_j(x)\le 0\},\: 1\le j\le m,
\end{equation}
which hereinafter will be called the feasible domain for this constraint. The feasible domain $Q$ of initial problem \eqref{eq4:problem} can be defined by the condition
\begin{equation}
  \label{eq4:Q_union}
  Q=Q_1\cap\dots\cap Q_m.
\end{equation}

@@

\begin{table}
  \caption{The number of problems with a small measure of the interval $[\alpha,\beta]$}
\begin{center}
  \label{tab4:exp_results2}
  \begin{tabular}{cccccccccc}
    \hline
     & 0.01\% & 0.02\% & 0.03\% & 0.04\% & 0.05\% & 0.06\% & 0.07\% & 0.08\% & 0.09\% \\ \hline
    $G_{SH}$ & 0 & 1 & 2 & 1 & 0 & 0 & 1 & 1 & 1  \\ 
    $G_{HL}$ & 3 & 6 & 2 & 3 & 5 & 1 & 2 & 1 & 2 \\ \hline
  \end{tabular}
\end{center}
\end{table}

@@

The first $p$ trial points are selected at the internal points of the interval $[a,b]$ arbitrarily. Each processor starts the trials execution at the corresponding points. After completing the $k$-th trial, when the points $x^{k+1},\dots, x^{k+p - 1}$ have already been selected but the trial results at these points are not obtained yet, the point of the next trial $x^{k+p}$ is selected according to the following rules.

@@

\subsection{The convergence conditions of the algorithm}
\begin{theorem}
  \label{th4:convergence_par}
  Assume problem \eqref{eq4:problem} to have a solution $x^*$ and the following conditions to be fulfilled:
  \begin{itemize}
    \item each subdomain $Q_j ,1\le j\le m+1$ from \eqref{eq4:q} is a union of a finite number of positive length segments;
    \item each function $g_j ,1\le j\le m+1$ satisfies Lipschitz condition with the corresponding constant $L_j$;
    \item the components $\epsilon_\nu$ of the reserve vector $\epsilon_R$ satisfy the inequalities
    \begin{equation}
      \label{eq4:reserves_par}
      0<2\epsilon_\nu<L_\nu(\beta-\alpha)
    \end{equation}
    where $\beta-\alpha$ is the length of the interval $[\alpha,\beta]$ belonging to the feasible domain $Q$ from \eqref{eq4:Q_union} and including the point $x^*$ which is the solution of the problem;
    \item beginning from sufficiently large step number $k$, the values $\nu_\nu$ from \eqref{eq4:lip_est_par} corresponding to the nonempty sets $I_\nu$ satisfy the inequalities
    \begin{equation}
      r_\nu\mu_\nu>2L_\nu,\:1\le\nu\le m+1,
    \end{equation}
    where the algorithm parameters $r_\nu>1,\:1\le\nu\le m+1$;
    \item during the parallelization, the inequalities
    \begin{displaymath}
      T^j_{l+1}-T^j_j<T<\infty,1\le j\le p, l>1,
    \end{displaymath}
    are true where $T_l^j$ is the moment of completing the $l$-th trial by the $j$-th processor and the inequality $1<p<P<\infty$ takes place for the number of processors $p$.
  \end{itemize}
  Then:
  \begin{itemize}
    \item the point $x^*$ is a limit one for the sequence ${x^k }$ generated by the described asynchronous algorithm for problem \eqref{eq4:problem} if $\epsilon=0$ in termination condition \eqref{eq4:stop_cond_par};
    \item any limit point $\overline x$ of the sequence $\{x_k \}$ is a solution of the problem \eqref{eq4:problem};
    \item the convergence to the limit point $\overline x$ is bilateral if $\overline x\not=a$ and $\overline x\not=b$.
  \end{itemize}
\end{theorem}

@@

\begin{proof}
  Since $\overline x$ is a limit point, there exists a sequence of the trial numbers $q^1,q^2,\dots$ satisfying the conditions
  \begin{equation}
    x^{q+1},\overline x\in[x_{t-1},x_t], t=t(q),q\in\{q^p\},p=1,2,\dots.
  \end{equation}
  These conditions reflect the fact that the point of the $(q+1)$-th trial falls into the interval $[x_{t-1},x_t]$ containing the point $\overline x$ at the step $q$, i.e., $t=t(q)$.

@@

  At every step $k\ge 1$ there exists an interval $(x_{i-1},x_i), \:i=i(k)$, at one of the ends of which the value $z^*_\omega$ from \eqref{eq4:z_const_par} is achieved. This value corresponds to the index $\omega$ defined (when $\omega<m$) by the conditions $I_\omega\not= \varnothing,\: I_{\omega+1} =  \varnothing$. If $I_{m+1}= \varnothing$, the index $\omega=m+1$. If at that $\nu(x_{i -1} )\not = \nu(x_i )$, then according (4.47) and (4.50), $R(i)=2\Delta_i>0$. In the case when $\nu(x_{i -1} )= \nu(x_i )=\omega$, the estimate
  \begin{displaymath}
    \max(z_{i-1}, z_i)-z^*_\omega\le\mu\Delta_i
  \end{displaymath}
  follows from \eqref{eq4:lip_est_par} and, along with \eqref{eq4:characteristic_par}, gives the estimate
  \begin{displaymath}
    R(i)=\Delta_i\left(1-\frac{\max(z_{i-1},z_i)-z^*_\omega}{r_\omega\mu_\omega\Delta_i}\right)\ge (1-r^{-1})\Delta_i,
  \end{displaymath}
  where $r$ is from (4.61). Thus, at any search step $k\ge 1$ there exists an interval having a strictly positive characteristic. Hence, taking into account the rules of selection of the interval with the greatest characteristic \eqref{eq4:MaxR_par}, one can conclude that the intervals from system (4.54) should also have the positive characteristics since these ones undergo splitting up by the trial points falling into these intervals. Consequently, conditions (4.57) hold.
